---
title: "STA305/1004 - Review Class"
fontsize: 9pt
date: "February 27, 2017"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
---

# Today's class

- Review for midterm test


# Midterm Test Information

**Date:** Wednesday, March 1

**Time:** The test will begin at approximately 11:10.  The time allowed to complete the test will be 90 minutes.

**Location:** 

- If your last name begins with a letter between A-M then you will write in AH400 (our classroom).
- If your last name begins with a letter between N-Z then you will write in EX310.

**Test Preview:**

Available at 1:00PM today on Portal.



# Q1

Consider a randomized study of two medical treatments A and B.  Three subjects are randomized to treatment A and three subjects are randomized to treatment B.  The response measured is mortality, $y$, after 6-weeks on treatment. Primary question: is there a difference in average mortality between the two treatments?

```{r,cache=TRUE,echo=FALSE}
set.seed(12)
yA <- round(rnorm(3,10,3),2)
yB <- round(rnorm(3,12,4),2)
```


The data are below:

Subject |$y$ | Treatment
--|--|--
1|`r yA[1]` |  A
5|`r yA[2]` |  A
4|`r yA[3]` |  A
6|`r yB[1]` |B
3|`r yB[2]` |B
2|`r yB[3]` |B

(a) Is the treatment assignment ignorable?
(b) How could subjects be randomized to treatment?

# Q1

```{r,comment=""}
y <- c(yA,yB);observed <- mean(yA)-mean(yB);observed #obs mean diff
index <- combn(1:6,3); res<-numeric(20)
for(i in 1:20){res[i] <- mean(y[index[,i]])-mean(y[-index[,i]])}
round(res,2)
mean(res)
round(res-mean(res),2)
round(observed-mean(res),2)
#sum(abs(round(res-mean(res),2))>=round(observed-mean(res),2))/20
#hist(res)
```

# Q1

(c) Calculate the randomization P-value?
(d) Does the treatment *cause* a change in average mortality?

# Q2

(a)  What is statistical power?
(b)  Suppose that a study is designed to test $H_0: \mu=0$ vs. $H_1:\mu<0$.  The study is conducted with $n=10, \sigma=1$ using $\alpha=0.05$.  The data is analyzed: ${\bar x}=-0.36,P\text{-value}=0.13$.  Is the reason that the study is not significant due to low power?  



