---
title: "STA305/1004 - Review of Statistical Theory" 
date: "January 11, 2017"
output: 
  beamer_presentation: 
    fig_width: 8
    fig_height: 7
    keep_tex: yes
fontsize: 9pt
---

# Data

Experimental data describes the outcome of the experimental run. For example 10 successive runs in a chemical experiment produce the following data: 

```{r,cache=TRUE}
set.seed(100)
# Generate a random sample of 5 observations 
# from a N(60,10^2)
dat <- round(rnorm(5,mean = 60,sd = 10),1) 
dat
```

# Distributions

Distributions can be displayed graphically or numerically.  

A histogram is a graphical summary of a data set.

```{r}
summary(dat)
```

# Distributions

```{r}
hist(dat)
```

***

# Distributions

- The total aggregate of observations that might occur as a result of repeatedly performing a particular operation is called a **population** of observations.
- The observations that actually occur are a **sample** from the population.

# Continuous Distributions

- A continuous random variable $X$ is fully characterized by it's density function $f(x)$. 
- $f(x) \ge 0$, $\thinspace f$ is piecewise continuous, and $\int_{-\infty}^{\infty}f(x)dx=1$.  

- The cumulative distribution function (CDF) of $X$ is defined as:

$$ F(x)=P(X \le x)=\int_{-\infty}^{x}f(x)dx.$$

# Continuous Distributions

- If $f$ is continuous at $x$ then $F'(x)=f(x)$ (fundamental theorem of calculus).  
- The CDF can be used to calculate the probability that $X$ falls in the interval $(a,b)$.  This is the area under the density curve which can also be expressed in terms of the CDF:

$$P\left(a < X < b\right)=\int_{a}^{b}f(x)dx=F(b)-F(a).$$

- In R a list of all the common distributions can be obtained by the command `help("distributions")`.
- For example, the normal density and CDF are given by `dnorm()` and `pnorm()`.

# Continuous Distributions

100 observations (using `rchisq()`) from a Chi-square distribution on 10 degrees of freedom $\chi^2_{10}$.  The density function of the $\chi^2_{10}$ is superimposed over the histogram of the sample.

```{r, echo=FALSE}
x<- rchisq(100,10) # draw a sample of 100 from chi-square 10
h <- hist(x) # create the histogram
# superimpoise chi-square density over histogram
xfit<-seq(min(x),max(x),length=40)
yfit <- dchisq(xfit,10) #chi-square density
yfit <- yfit*diff(h$mids[1:2])*length(x)
lines(xfit,yfit)
```


# Randomness

- A random drawing is where each member of the population has an equal chance of being selected. 
- The hypothesis of random sampling may not apply to real data.
- For example, cold days are usually followed by cold days.   
- So daily temperature not directly representable by random drawings. 
- In many cases we can’t rely on the random sampling property although design can make this assumption relevant. 

# Parameters and Statistics

What is the difference between a parameter and a statistic?  

- A parameter is a population quantity and a statistic is a quantity based on a sample drawn from the population.  

Example: The population of all adult (18+ years old) males in Toronto, Canada.  

- Suppose that there are $N$ adult males and the quantity of interest, $y$, is age.  
- A sample of size $n$ is drawn from this population.  
- The population mean is $\mu=\sum_{i=1}^N y_i /N$.
- The sample mean is ${\bar y}=\sum_{i=1}^n y_i /n$.

# Residuals and Degress of Freedom

$y_i-{\bar y}$ is called a residual.  

- Since $\sum (y_i-{\bar y})=0$ any $n-1$ completely determine the the last observation.  
- This is a constraint on the the residuals.  
- So $n$ residuals have $n-1$ degrees of freedom since the last residual cannot be freely chosen.  

# The Normal Distribution

The density function of the normal distribution with mean $\mu$ and standard deviation $\sigma$ is:

$$ \phi(x)=\frac{1}{\sigma \sqrt{2\pi}}exp\left( \frac{-1}{2} \left(\frac{x-\mu}{\sigma}\right)^2\right)$$

The cumulative distribution function (CDF) of a $N(0,1)$ distribution,

$$ \Phi(x)= P(X<x)= \int_{-\infty}^x \phi(x)dx$$

# The Normal Distribution
```{r}
x <- seq(-4,4,by=0.1)
plot(x,dnorm(x),type="l",main = "The Standard Normal Distribution", 
     ylab=expression(paste(phi(x))))
```

***

# The Normal Distribution

```{r}
plot(x <- seq(-2,2,by=0.1),pnorm(x),type="l",
     xlab="x",ylab=expression(paste(Phi(x))),
     main = "Standard Normal Cumulative Distribution Function")
```

***

# The Normal Distribution

A random variable $X$ that follows a normal distribution with mean $\mu$ and variance $\sigma^2$ will be denoted by $$X \sim N\left(\mu, \sigma^2\right).$$

If $Y \sim N\left(\mu, \sigma^2\right)$ then $$Z \sim N(0,1),$$ where $$Z=\frac{Y-\mu}{\sigma}.$$  

***

# The Normal Distribution

$X \sim N(5,3)$.  Use R to find $P(4<X<6)$.

```{r}
pnorm(6,mean = 5,sd = sqrt(3))-pnorm(4,mean = 5,sd = sqrt(3))
```

# Normal Quantile Plots

The following data are the weights from 11 tomato plants.

```{r,echo=FALSE,cache=TRUE,warning=FALSE}
library(BHH2); 
data(tomato.data)
tomato.data$pounds
```

Do the weights follow a Normal distribution?


# Normal Quantile Plots

A normal quantile plot in R can be obtained using `qqnorm()` for the normal probability plot and `qqline()` to add  the straight line.

```{r}
qqnorm(tomato.data$pounds); qqline(tomato.data$pounds)
```


# Central Limit Theorem

The central limit theorem states that if $X_1, X_2, ...$ is an independent sequence of identically distributed random variables with mean $\mu=E(X_i)$ and variance $\sigma^2=Var(X_i)$ then

$$ \lim_{n\to\infty} P\left(\frac{\bar X - \mu}{\frac {\sigma}{\sqrt n}} \leq x \right) = \Phi(x),$$

where ${\bar X} = \sum_{i=1}^{n} X_i/n$ and $\Phi(x)$ is the standard normal CDF.  This means that the distribution of ${\bar X}$ is approximately $N\left(\mu,{\frac {\sigma}{\sqrt n}}\right)$.

# Central Limit Theorem

Example: A fair coin is flipped 50 times.  What is the distribution of the average number of heads?


# Central Limit Theorem

```{r}
set.seed(100)
Total.heads <- rbinom(100,50,0.5); Ave.heads <- Total.heads/50; 
hist(Ave.heads)
```

***

# Central Limit Theorem

```{r,echo=FALSE}
set.seed(100)
x<- rbinom(100,50,0.5)/50 # draw a sample of 100 from chi-square 10
h <- hist(x) # create the histogram
# superimpoise chi-square density over histogram
xfit<-seq(min(x),max(x),length=40)
yfit <- dnorm(xfit,mean = .5,sd = sqrt((.5*.5)/50)) 
yfit <- yfit*diff(h$mids[1:2])*length(x)
lines(xfit,yfit)
```

***

# Chi-Square Distribution

Let $X_1, X_2, ..., X_n$ be independent and identically distributed random variables that have a $N(0,1)$ distribution. The distribution of 

$$ \sum_{i=1}^{n}X_i^2,$$ 

has a chi-square distribution on $n$ degrees of freedom or $\chi^2_{n}$.

The mean of a $\chi^2_{n}$ is $n$ with variance $2n$.

# Chi-Square Distribution

Let $X_1,X_2,...,X_n$ be independent with a $N(\mu, \sigma^2)$ distribution. What is the distribution of the sample variance $S^2=\sum_{i=1}^{n}(X_i-{\bar X})^2/(n-1)$?

# t Distribution

If $X \sim N(0,1)$ and $W \sim \chi^2_n$ then the distribution of $\frac{X}{\sqrt{W/n}}$ has a t distribution on $n$ degrees of freedom or $\frac{X}{\sqrt{W/{n}}} \sim t_n$.

# t Distribution

Let $X_1, X_2, ...$ is an independent sequence of identically distributed random variables that have a $N(0,1)$ distribution. What is the distribution of 

$$ \frac{{\bar X}-\mu }{\frac {S}{\sqrt {n-1}}}$$ 

where $S^2=\sum_{i=1}^{n}(X_i-{\bar X})^2/(n-1)$?  


# t Distribution

```{r,echo=FALSE}
# Compare t densities with normal density
x <- seq(-4, 4, length=100)
hx <- dnorm(x) #normal density

degf <- c(1, 3, 8, 30)
colors <- c("red", "blue", "darkgreen", "gold", "black")
labels <- c("df=1", "df=3", "df=8", "df=30", "normal")

plot(x, hx, type="l", lty=2, xlab="x value",
     ylab="Density", main="Comparison of t Distributions")

for (i in 1:4){
  lines(x, dt(x,degf[i]), lwd=2, col=colors[i]) # dt is the t density
}

legend("topright", inset=.05, title="Distributions",
       labels, lwd=2, lty=c(1, 1, 1, 1, 2), col=colors)
```


***


# F Distribution

Let $X \sim \chi^2_m$ and $Y\sim \chi^2_n$ be independent. The distribution of 

$$ W= \frac{X/m}{Y/n} \sim F_{m,n},$$

where $F_{m,n}$ denotes the F distribution on $m,n$ degrees of freedom. The F distribution is right skewed (see graph below).  For $n>2, E(W)=n/(n-2)$. It also follows that the square of a $t_n$ random variable follows an $F_{1,n}$. 

***


# F Distribution

```{r, echo=FALSE}
# Compare t densities with normal density

colors <- c("red", "blue")
plot(seq(0,5,by=0.1), df(seq(0,5,by=0.1),7,10),type="l",col=colors[1],ylab="Density",xlab="x",main = "F Distributions")
lines(seq(0,5,by=0.1), df(seq(0,5,by=0.1),7,4),type="l",lty=2,col=colors[2])
labels <- c("F(7,10)","F(7,4)")
legend("topright", inset=.05, title="Distributions",
       labels, lwd=2, lty=c(1, 2), col=colors)
```


***

# Linear Regression

Lea (1965) discussed the relationship between mean annual temperature and mortality index for a type of breast cancer in women taken from regions in Europe (example from Wu and Hammada).

The data is shown below.

```{r,cache=TRUE}
#Breast Cancer data
M <- c(102.5, 104.5, 100.4, 95.9, 87.0, 95.0, 88.6, 89.2, 
       78.9, 84.6, 81.7, 72.2, 65.1, 68.1, 67.3, 52.5)
T <- c(51.3, 49.9, 50.0,49.2, 48.5, 47.8, 47.3, 45.1, 
       46.3, 42.1, 44.2, 43.5, 42.3, 40.2, 31.8, 34.0)
```

# Linear Regression

A linear regression model of mortality versus temperature is obtained by estimating the intercept and slope in the equation:

$$ y_i = \beta_0 + \beta_1 x_i + \epsilon_i, i=1,...,n$$

where $\epsilon_i \sim N(0,\sigma^2)$.  The values of $\beta_0, \beta_1$ that minimize the sum of squares

$$ \sum_{i=1}^{n} (y_i-(\beta_0+\beta_1 x_i))^2, $$

are called the least squares estimators. They are given by: 

- $\hat{\beta_0}={\bar y}-{\hat{\beta_1}}{\bar{x}}$ 
- ${\hat{\beta_1}}=r\frac{S_y}{S_x}$ 

$r$ is the correlation between $y$ and $x$, and $S_x, S_y$ are the sample standard deviations of $x$ and $y$ respectively.


***

# Linear Regression

```{r}
plot(T,M,xlab="temperature",ylab="mortality index")
```

***

# Linear Regression

```{r,cache=TRUE}
reg1 <- lm(M~T)
summary(reg1) # Parameter estimates and ANOVA table
```

***
# Linear Regression

```{r}
plot(T,M,xlab="temperature",ylab="mortality index")
abline(reg1) # Add regression line to the plot
```


# Linear Regression
```{r}
#plot residuals vs. fitted
plot(reg1$fitted,reg1$residuals);
abline(h=0) # add horizontal line at 0
```


***


# Linear Regression
```{r}
#check normality of residuals
qqnorm(reg1$residuals); qqline(reg1$residuals)
```


***

# Linear Regression

If there is more than one independent variable then the above model is called a multiple linear regression model.

$$ y_i=\beta_0+\beta_1 x_{i1} + \beta_2 x_{i2} + \cdots +  \beta_k x_{ik} + \epsilon_{i}, \thinspace i=1,...,n, $$

where $\epsilon_{i} \sim N(0,\sigma^2)$.

This can also be expressed in matrix notation as

$$ y=X\beta+\epsilon$$

The least squares estimator is
 
 $${\hat \beta}=\left(X^{T} X\right)^{-1}X^{T}y.$$

The covariance matrix of ${\hat \beta}$ is $\left(X^{T} X\right)^{-1}{\sigma}^2$.  An estimator of $\sigma^2$ is 

$${\hat \sigma^2}=\frac{1}{n-k}\sum_{i=1}^{n}(y_i-{\hat y_i})^2,$$

where ${\hat y_i}={\hat \beta_0}+{\hat \beta_1} x_{i1} + \cdots +  {\hat \beta_k} x_{ik}$ is the predicted value of $y_i$.

# Weighing Problem

Harold Hotelling in 1949 wrote a paper on how to obtain more accurate weighings through experimental design.

**Method 1**

Weigh each apple separately. 

**Method 2**

Obtain two weighings by 

1. Weighing two apples in one pan.
2. Weighing one apple in one pan and the other apple in the other pan 

# Weighing Problem

Let $w_1, w_2$ be the weights of apples one and two.  Each weighing has standard error $\sigma$. So the precision of the estimates from method 1 is $\sigma$.

If the objects are weighed together in one pan, resulting in measurement $m_1$, then in opposite pans, resulting in measurement $m_2$, we have two equations for the unknown weights $w_1,w_2$:

$$w_1+w_2 =  m_1$$ 
$$w_1-w_2 =  m_2$$


# Weighing Problem

This can also be viewed as a linear regression problem $y=X{\beta}+\epsilon$:

$$y=(m_1,m_2)^{\prime},\thinspace X=
 \begin{pmatrix}
  1 & 1 \\
  1 & -1 \\
 \end{pmatrix}, \thinspace {\beta}=(w_1,w_2)^{\prime}.$$

# Weighing Problem
 
 The least-squares estimates can be found using R.

```{r}
#step-by-step matrix mutiplication example for weighing problem

X <- matrix(c(1,1,1,-1),nrow=2,ncol=2) #define X matrix
Y <- t(X)%*%X # multiply X^T by X (X^T*X) NB: t(X) is transpose of X
W <- solve(Y) # calculate the inverse
W %*% t(X) # calculate (X^T*X)^(-1)*X^T
W # print (X^T*X)^(-1) for SE 
```



