---
title: "STA305/1004-Class 16"
output: beamer_presentation
fontsize: 8pt
date: "March 13, 2017"
---

```{r,echo=FALSE,cache=TRUE,warning=FALSE}
tab0401 <- read.table("~/Dropbox/Docs/sta305/BHHData/BHH2-Data/tab0401.dat", header=TRUE, quote="\"")
library(BHH2)
```


```{r,cache=TRUE,comment="",cache=TRUE,echo=FALSE}
#contrasts(tab0401$diets)
lm.diets <- lm(y~diets,data=tab0401)
#round(summary(lm.diets)$coefficients,2)
```


```{r,echo=F,cache=T,warning=FALSE}
library(tidyr)
data_wide <- spread(tab0401, diets, y)
A <- data_wide$A[is.na(data_wide$A)==F]
B <- data_wide$B[is.na(data_wide$B)==F]
C <- data_wide$C[is.na(data_wide$C)==F]
D <- data_wide$D[is.na(data_wide$D)==F]
ave <- c(mean(A),mean(B),mean(C),mean(D))
gave <- rep(mean(tab0401$y),4)
diff <- ave-gave
dat <- rbind(cbind(A,B,C,D),ave,gave,diff)
row.names(dat) <- c("","","","","","","Treatment Average", "Grand Average","Difference")
#knitr::kable(dat,col.names=c("A","B","C","D"))
```


# Today's Class

- Sample size for ANOVA
- Factorial designs at two levels
- Cube plots
- Calculation of factorial effects



# Sample size for ANOVA - Designing a study to compare more than two treatments 

- Consider the hypothesis that k means are equal vs. the alternative that at least two differ. 
- What is the probability that the test rejects if at least two means differ? 
- Power = $1-P({\text{Type II error}})$ is this probability.

# Sample size for ANOVA - Designing a study to compare more than two treatments 

The null and alternative hypotheses are:

$$H_0: \mu_1=\mu_2 = \cdots = \mu_k \thinspace {\text  vs. } \thinspace H_1: \mu_i \ne\mu_j.$$ 

The test rejects at level $\alpha$ if

$$MS_{Treat}/MS_E \ge F_{k-1,N-K,\alpha}.$$

The power of the test is

$$ 1- \beta= P\left(MS_{Treat}/MS_E \ge F_{k-1,N-K,\alpha} \right),$$

when $H_0$ is false.  

# Sample size for ANOVA - Designing a study to compare more than two treatments 

When $H_0$ is false it can be shown that:

- $MS_{Treat}/\sigma^2$ has a non-central Chi-square distribution with $k-1$ degrees of freedom and non-centrality parameter $\delta$.

- $MS_{Treat}/MS_E$ has a non-central $F$ distribution with the numerator and denominator degrees of freedom $k-1$ and $N-k$ respectively, and non-centrality parameter 

$$\delta = \frac{\sum_{i=1}^kn_i\left(\mu_i-{\bar \mu} \right)^2}{\sigma^2},$$

where $n_i$ is the number of observations in group $i$, ${\bar \mu}=\sum_{i=1}^k \mu_i/k$, and $\sigma^2$ is the within group error variance .

This is dentoted by $F_{k-1,N-k}(\delta)$.

# Direct calculation of Power

- The power of the test is 

$$P\left(F_{k-1,N-k}(\delta) > F_{k-1,N-K,\alpha} \right).$$

- The power is an increasing function $\delta$ 
- The power depends on the true values of the treatment means $\mu_i$, the error variance $\sigma^2$, and sample size $n_i$.
- If the experimentor has some prior idea about the treament means and error variance the sample size (number of replications) that will guaruntee a pre-assigned power of the test.

# Blood coagulation example - sample size

Suppose that an investigator would like to replcate the blood coagulation study with only 3 animals per diet.  In this case $k=4, n_i=3.$  The treatment means from the initial study are:

Diet | A | B  | C | D 
-----|---|---|---|---
Average | `r ave[1]` | `r ave[2]` | `r ave[3]`| `r ave[4]`|

```{r}
anova(lm.diets)
```

# Blood coagulation example - sample size

- $\mu_1=$ `r ave[1]`, $\mu_2=$ `r ave[2]`, $\mu_3=$ `r ave[3]`, $\mu_4=$ `r ave[4]`.  
- The numerator df = 4-1=3, and the denominator df = 12-4=8.
- The error variance $\sigma^2$ was estimated as $MS_E=5.6$.  
- Assuming that the estimated values are the true values of the parameters, the non-centrality parameter of the $F$ distribution is

$$\delta = 3 \times \left((61-64)^2+(66-64)^2+(68-64)^2+(61-64)^2\right)/5.6=20.35714$$

# Blood coagulation example - sample size

If we choose $\alpha=0.05$ as the significance level then $F_{3,20,0.05}=$ `r qf(p = .05,df1 = 3,df2 = 8,lower.tail=F)`. The power of the test is then 

$$P\left(F_{3,8}(20.35714) > 4.066181 \right)=0.8499.$$

This was calculated using the CDF for the $F$ distribution in R `pf()`.

```{r}
1-pf(q = 4.066181,df1 = 3,df2 = 8,ncp = 20.35714)
```

# Calculating power and sample size using the pwr library

- There are several libraries in R which can calculate power and sample size for statistical tests.  The library `pwr()` has a function for ANOVA.

- `pwr.anova.test(k = NULL, n = NULL, f = NULL, sig.level = 0.05, power = NULL)` 

for computing power and sample size.

- `k`	Number of groups
- `n`	Number of observations (per group)
- `f` Effect size

The effect size $f$ 

$$f = \sqrt{\frac{\sum_{i=1}^k\left(\mu_i-{\bar \mu} \right)^2/k}{\sigma^2}},$$
is related to the non-centrality parameter $\delta$ via $\delta=k\cdot n_i\cdot f^2$.

- $n_i$ is the number of observations in group $i$, ${\bar \mu}=\sum_{i=1}^k \mu_i/k$, and $\sigma^2$ is the within group error variance.

# Calculating power and sample size using the pwr library

In the previous example $\delta=20.35714$ so $f=\sqrt{\frac{\delta}{k \cdot n_i}}=\sqrt{20.35714/4\cdot 3}=$ `r sqrt(20.35714/(12))`.

```{r,warning=FALSE}
library(pwr)
pwr.anova.test(k = 4,n = 3,f = 1.30247)
```

# Calculating power and sample size using the pwr library

```{r,echo=FALSE}
library(pwr)
x <- seq(.05,5,by=0.01)
plot(x,pwr.anova.test(k = 4,n = 3,f = x)$power,type="l",
     xlab="Effect Size",ylab="Power",main="Power vs. Effect Size for k=4, n=3")
```



